Investigating the Applications and Limits of Single Asymmetric Particle Light Scattering using Limited Detection Schemes

### 2022-11-14 Manuscript.tex
fixing notation

To demonstrate the analysis, we study a simple paradigmatic example of a trapped entity more complex than a single sphere, i.e. a trapped asymmetric dimer; and we explore how to optimise the experimental measurement to obtain the best estimate of the dimer's instantaneous orientation, as a further paradigmatic example of interpretation of the scattering data to obtain dynamical information on the trapped entity. There is a paucity of literature on measuring the orientation of trapped non-spherical particles till the recent attempt in \cite{raudsepp2022estimating} where imaging was employed to study the orientation of trapped dimers and a resolution of () was achieved. Here, we simulate the Brownian motion of a dimer and calculate the resulting scattered light, and show how Bayesian inference can be used to optimise extraction of the true dimer orientation from the light scattering signals given a choice of scattering measurement angles.

\begin{figure}
	\centering
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1a.png}
		\subcaption{}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1b.png}
		\subcaption{}
	\end{subfigure}
	\caption{Experimental set up, a) Side view, b) Top view }
	\label{fig:setup}
\end{figure}

@@

We represent the instantaneous orientation of the moving dimer in the
trap as a three component unit vector, $\hat{\bf s}$, pointing from
the centre of the larger sphere to the centre of the smaller sphere.
A plane wave ``probe'' laser beam directed perpendicular to the
trapping laser, incident on the dimer, generates a scattering pattern
dependent on the dimer's instantaneous orientation,
$I(\bar{\bf s},\ \theta)$. This scattering pattern was computed using MSTM
for each time step of our Brownian simulation.  To represent a
simulated 'measurement', three angles ($\theta_1$, $\theta_2$,
$\theta_3$) were chosen and the calculated intensity at each angle k
recorded as $I(\hat{\bf s}, \theta_k)$.

Our goal is to estimate the dimer orientation $\bar{\bf s}$ over time,
based on the scattering measurements $I(\bar{\bf s},\ \theta_k)$: the
question then is what set of angles ($\theta_1$, $\theta_2$,
$\theta_3$) give us the most useful information on $\hat{\bf s}$.  To
initially simplify the problem, rather than trying to get an exact
estimation of $\hat{\bf s}$ we discretise the orientation space into
$30$ possible reference orientations $\hat{\bf n}_i$ that divide up a
unit sphere evenly \cite{Rey2006}: our question becomes, how can we
optimise the experiment to identify the reference orientation
$\hat{\bf n}_i$ closest to the real orientation $\hat{\bf s}$.
(Ultimately our method could be extended to an arbitrarily large
number of reference orientations but for clarity and practicability we
use a relatively small number here; additionally, the choice of a
small number of discrete reference or 'candidate' orientations also
accounts for the fact that any experiment will necessarily have a
limited precision.)  Using MSTM we compute the scattering that would
be produced by the dimer if it were in each of the reference
orientations, denoted as $I(\hat{\bf n}_i, \ \theta)$ and obtain
'measurements' $I(\hat{\bf n}_i, \ \theta_k)$ at our chosen set of angles
$\theta_k$. These raw intensities are normalised according to
\eqref{eq:scale}:

\begin{align}
	\label{eq:scale}
	y_k(\hat{\bf n}_i,\ \theta_k) = 
	 \frac{\left[I(\hat{\bf n}_i, \ \theta_k)- \langle I(\hat{\bf n}, \ \theta_k) \rangle \right]} 
	{\langle I^2(\hat{\bf n},\ \theta_k) \rangle -\langle I(\hat{\bf n}, \ \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values
of $I(\hat{\bf n}, \theta_k)$.  The dimer's \emph{actual}
instantaneous scattered intensity based on the real instantaneous
orientation $\bar{\bf s}$ is similarly scaled using \eqref{eq:scale}
to give $y(\bar{\bf s},\ \theta_k)$.  See Tables A.1 and \ref{tab:A2}
for numerical values of our reference orientation, raw intensities,
and scaled signals.  In practice, $y(\bar{\bf s}\, \theta_k)$ will
have an inherent amount of uncertainty in our measurement due to noise
from our detector.  We compare our instantaneous signal to our
reference signal by assuming that the measurement uncertainty for each
detection fibre is Gaussian and is centred about $y_k(\bar{\bf s})$
Combining all three detection angles provides us with the conditional
probability that, should the dimer be in orientation $\hat{\bf n}_i$,
it would produce a signal $y(\bar{\bf s})$:
\begin{align}
	\label{eq:conditional}
	p_i(y(\bar{\bf s})|\hat{\bf n}_i) = \Pi^3_{k=1}
	(2\pi\sigma_k^2)^{-1/2} 
	e^{-(y_{k}(\hat{\bf n}_i)-y_{k}(\bar{\bf s}))^2/2\sigma_k^2}
\end{align}
where $\sigma_k$ is the average error in the detected signal from our
dimer $y_k(\bar{\bf s})$.  We left this as a tuneable variable to
measure how the influence of experimental error would effect our
predictive model.  The signal error was simulated in our model by
stochastically adjusting our values of $y_k(\bar{\bf s})$.
\begin{eqnarray*}
  \sigma_k
  &= \epsilon\bar{y}(\hat{\bf n}_k)
  \\
  \Rightarrow y(\bar{\bf s}_k) &= y(\bar{\bf s})_k \pm n\sigma_k
  \\ 
n \in \mathbb{R}, n\in[0,1]
\end{eqnarray*} 
where $\epsilon$ is the percent error expected from our detection
fibres.  Equation~\eqref{eq:conditional} assigns a probability to each
reference orientation based on how similar the 'reference orientation
signal' $y_{k}(\hat{\bf n}_i)$ is to the actual signal
$y_{k}(\bar{\bf s})$ generated by the dimer in its real orientation
$\bar{\bf s}$.  However, if our values of $y_k(\hat{\bf n}_i)$ are
similar to one another the probability distribution from
Eq.~\eqref{eq:conditional} will therefore assign similar
probability for multiple reference orientations. Bayesian inference
allows one to update a probability distribution by accounting for
information known prior to making a measurement.  We apply this by
accounting for our distribution of reference orientations and
instantaneous signals.
\begin{align}
  \label{eq:bayes}
  p_i(\hat{\bf n}| y(\bar{\bf s}))
  &=
    \frac{p_i(y(\bar{\bf s})|\hat{\bf n})p_i(\hat{\bf n})}{p(y)}
\end{align}
where $p_i(\hat{\bf n})$ and $p(y)$ are our estimations of the prior
distributions of particle orientations and instantaneous signals
respectively.  The former can be estimated by some Boltzmann
distribution based on how far apart reference orientations are in
physical space, covered in Section~\ref{sec:2.3}.  The latter prior
$p(y)$ is the probability of our dimer producing a signal
$y_k(\bar{\bf s})$ that falls with our distribution of reference
signals $y_k(\hat{\bf n})$.  This is given by taking the discrete
integral over collection of reference orientations.
\begin{align}
  p(y)
  =
  \sum_{i=1}^{30} p_i(y(\bar{\bf s})|\hat{\bf n})p_i(\hat{\bf n})
\end{align}
The implementation of Eq.~\eqref{eq:bayes} improves our estimate by
suppressing reference orientations that are far away from our dimer's
orientation vector in physical space.  In order to evaluate the
accuracy of our method, we tested it using a Brownian dynamics
simulation of the dimer (see Section~\ref{sec:2.1}).

>>>>

To demonstrate the analysis, we study a simple, illustrative example
of a trapped entity more complex than a single sphere (i.e.\ a trapped
asymmetric dimer).  We explore how to optimise the experimental
measurement to obtain the best estimate of the dimer's instantaneous
orientation, as a further paradigmatic example of interpretation of
the scattering data to obtain dynamical information on the trapped
entity.  There is a paucity of literature on measuring the orientation
of trapped non-spherical particles till the recent attempt in
Ref.~\citenum{raudsepp2022estimating} where imaging was employed to
study the orientation of trapped dimers and a resolution of () was
achieved.  Here, we simulate the Brownian motion of a dimer and
calculate the resulting scattered light, and show how Bayesian
inference can be used to optimise extraction of the true dimer
orientation from the light scattering signals given a choice of
scattering measurement angles.

\begin{figure}
\centering
\begin{subfigure}{0.45\textwidth}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1a.png}
	\subcaption{}
\end{subfigure}
\begin{subfigure}{0.45\textwidth}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1b.png}
	\subcaption{}
\end{subfigure}
\caption{\label{fig:setup}
  %
  Experimental set up, a) Side view, b) Top view
%
}
\end{figure}

@@

We represent the instantaneous orientation of the moving dimer in the
trap as a unit vector $\hat{\bf s}$, pointing from the centre of the
larger sphere to the centre of the smaller sphere.  A plane wave
``probe'' laser beam directed perpendicular to the trapping laser,
incident on the dimer, generates a scattering pattern dependent on the
dimer's instantaneous orientation, $I(\bar{\bf s},\theta_k)$.  This
scattering pattern was computed using MSTM for each time step of the
Brownian dynamics simulation.  To represent a simulated measurement,
three angles ($\theta_1$, $\theta_2$, $\theta_3$) were chosen, and the
calculated intensity at each angle $\theta_k$ is recorded as
$I(\hat{\bf s}, \theta_k)$.

Our goal is to estimate the dimer orientation $\bar{\bf s}$ over time,
based on the scattering measurements $I(\bar{\bf s},\ \theta_k)$: the
question then is what set of angles ($\theta_1$, $\theta_2$,
$\theta_3$) give us the most useful information on $\hat{\bf s}$.  To
initially simplify the problem, rather than trying to get an exact
estimation of $\hat{\bf s}$, we discretise the orientation into
$n_{\rm ref}=30$ possible reference orientations $\hat{\bf n}_\alpha$
that are evenly distributed on the unit sphere \cite{Rey2006}.
%
We then consider how to optimise an experiment to identify the
reference orientation $\hat{\bf n}_\alpha$ that is closest to the
actual orientation $\hat{\bf s}$ of the dimer.
%
The method can be extended to an arbitrarily large number of reference
orientations, but for clarity and practicability we use a relatively
small number here; additionally, the choice of a small number of
discrete reference or ``candidate'' orientations also accounts for the
fact that any experiment will necessarily have a limited precision.
%
Using MSTM we compute the scattering that would be produced by the
dimer in each of the reference orientations, denoted as
$I(\hat{\bf n}_\alpha, \theta)$ and obtain ``measurements''
$I(\hat{\bf n}_\alpha, \theta_k)$ at our chosen set of angles
($\theta_1$, $\theta_2$, $\theta_3$).
%
These raw intensities are normalised according to
\begin{align}
\label{eq:scale}
  y_k(\hat{\bf n}_\alpha)
  = 
  \frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
  {\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values
of $I(\hat{\bf n}, \theta_k)$.  The dimer's \emph{actual}
instantaneous scattered intensity, based on the real instantaneous
orientation $\bar{\bf s}$, is similarly scaled using
Eq.~\eqref{eq:scale} to give $y(\bar{\bf s},\theta_k)$.
%
The reference orientations, raw intensities, and scaled signals are
given in Tables~\ref{tab:A1} and \ref{tab:A2}.
%
In practice, $y_k(\bar{\bf s})$ will have an inherent amount of
uncertainty in our measurement due to noise from our detector.  We
compare our instantaneous signal to our reference signal by assuming
that the measurement uncertainty for each detection fibre is Gaussian
and is centred about $y_k(\bar{\bf s})$ Combining all three detection
angles provides us with the conditional probability that, should the
dimer be in orientation $\hat{\bf n}_\alpha$, it would produce a
signal $y(\bar{\bf s})$:
\begin{align}
\label{eq:conditional}
  p(y(\bar{\bf s})|\hat{\bf n}_\alpha)
  = \prod^3_{k=1}
(2\pi\sigma_k^2)^{-1/2} 
e^{-(y_{k}(\hat{\bf n}_\alpha)-y_{k}(\bar{\bf s}))^2/2\sigma_k^2}
\end{align}
where $\sigma_k$ is related to the uncertainty in the detected signal
from our dimer $y_k(\bar{\bf s})$.  We left this as a tuneable
parameter to characterize the influence of experimental error on our
predictive method.  The signal error was simulated in our model by
stochastically adjusting our values of $y_k(\bar{\bf s})$.
\begin{eqnarray*}
  \sigma_k
  &= \epsilon\bar{y}(\hat{\bf n}_k)
  \\
  \Rightarrow y(\bar{\bf s}_k) &= y(\bar{\bf s})_k \pm n\sigma_k
  \\ 
n \in \mathbb{R}, n\in[0,1]
\end{eqnarray*} 
where $\epsilon$ is the percent error expected from our detection
fibres.  Equation~\eqref{eq:conditional} assigns a probability to each
reference orientation based on how similar the ``reference orientation
signal'' $y_{k}(\hat{\bf n}_\alpha)$ is to the actual signal
$y_{k}(\bar{\bf s})$ generated by the dimer in its real orientation
$\bar{\bf s}$.  However, if our values of $y_k(\hat{\bf n}_\alpha)$
are similar to one another, the probability distribution from
Eq.~\eqref{eq:conditional} will therefore assign similar probability
for multiple reference orientations. Bayesian inference allows one to
update a probability distribution by accounting for information known
prior to making a measurement.  We apply this by accounting for our
distribution of reference orientations and instantaneous signals.
\begin{align}
  \label{eq:bayes}
  p(\hat{\bf n}_\alpha| y_1,y_2,y_3)
  &=
    \frac{p(y_1,y_2,y_3|\hat{\bf n}_\alpha)
    p(\hat{\bf n}_\alpha)}{p(y_1, y_2, y_3)}
\end{align}
where $p(\hat{\bf n}_\alpha)$ and $p(y_1, y_2, y_3)$ are the prior
estimates of the distributions of particle orientations and
instantaneous signals, respectively.  The former can be estimated by
some Boltzmann distribution based on how far apart reference
orientations are in physical space, covered in Section~\ref{sec:2.3}.
The latter prior $p(y)$ is the probability of our dimer producing a
signal $y_k(\bar{\bf s})$ that falls with our distribution of
reference signals $y_k(\hat{\bf n})$.  This is given by taking the
discrete integral over collection of reference orientations.
\begin{align}
  p(y_1, y_2, y_3)
  =
  \sum_{\alpha=1}^{n_{\rm ref}}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  p(\hat{\bf n}_\alpha)
\end{align}
The implementation of Eq.~\eqref{eq:bayes} improves our estimate by
suppressing reference orientations that are far away from our dimer's
orientation vector in physical space.  In order to evaluate the
accuracy of our method, we tested it using a Brownian dynamics
simulation of the dimer (see Section~\ref{sec:2.1}).
